Weighted Model Counting with Conditional Weights

### 2020-08-07 paper/paper.tex
Conclusions + editing

\begin{abstract} % TODO
  AAAI creates proceedings, working notes, and technical reports directly from
  electronic source furnished by the authors. To ensure that all papers in the
  publication have a uniform appearance, authors must adhere to the following
  instructions.
\end{abstract}

@@

\emph{Weighted model counting} (WMC), i.e., an extension of model counting
($\#\SAT$) that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}
has emerged as the unifying inference mechanism across many (probabilistic)
domains. It has been applied to Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906} and probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, probabilistic
logic programming languages \cite{DBLP:journals/tplp/FierensBRSGTJR15} and
probabilistic programs with discrete probability distributions
\cite{DBLP:journals/corr/abs-2005-09089}. Over the last fifteen years, WMC has
been extended and generalised in many ways, e.g., to handle continuous
probability distributions \cite{DBLP:conf/ijcai/BellePB15}, first-order
probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, the idea of weights in WMC
has been generalised from non-negative real numbers to an arbitrary semiring,
making it applicable to domains such as parsing, dynamic programming, constraint
programming, databases, and networks \cite{DBLP:journals/japll/KimmigBR17}.

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This simplification is motivated by the fact that the
number of models scales exponentially with the number of atoms, so listing the
weight of every model is intractable. However, this also severely restricts what
probability distributions can be represented. A common way to overcome this
limitation is by adding more literals. While we show that this is always
possible, the rest of the paper demonstrates that it can be significantly more
efficient to encode weights in a more flexible way instead.

@@

Let $\mathcal{L}$ be a propositional logic with atoms $a$ and $b$, and let $U =
\{ a, b \}$. Then $2^U$, the power set of $U$, is the set of all models of
$\mathcal{L}$, and $2^{2^U}$ is the set of all formulas. These sets can also be
represented as Boolean algebras (e.g., $(2^{2^U}, \land, \lor, \neg, \bot,
\top)$) with a partial order $\le$ that corresponds to set inclusion
$\subseteq$---see \cref{tbl:notation_example} for examples of how various
elements can be represented in both notations. Most importantly, note that the
word \emph{atom} has completely different meanings in logic and in Boolean
algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an element of
$U$, whereas an atom in a Boolean algebra is (in set-theoretic terms) a
singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom while referring to
logical atoms as \emph{variables}. In the rest of the paper, for any set $U$, we
will use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

\subsection{The Space of Functions on Boolean Algebras}

We build on the definitions of multiplication and projection by
\citet{DBLP:conf/aaai/DudekPV20} and define more operations that can be used to
manipulate functions from Boolean algebras to non-negative real numbers. All of
these operations have efficient implementations in the CUDD
\cite{somenzi1998cudd} package for manipulating ADDs (among other things) that
is used by ADDMC {\cite{DBLP:conf/aaai/DudekPV20}.

@@

\section{WMC as a Measure} \label{sec:wmc_as_measure}

Let $U$ be a set. A \emph{measure} is a function $\mu\colon 2^{2^U} \to
\mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$, and $\mu(a \lor b) = \mu(a) +
\mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$. A \emph{weight
  function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$. A weight
function is \emph{factored} if $\nu = \prod_{x \in U} \nu_x$ for some functions
$\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that a weight
function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a measure
$\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if $\mu_\nu(x) = \sum_{\{u\} \le
  x} \nu(u)$.

@@

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a \emph{weight function} defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \begin{split}
    \mathrm{WMC}(\Delta) &= \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) \\
    &= w(a)w(b) + w(a)w(\neg b) = 0.3.
  \end{split}
\end{equation}
Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon 2^{\{b\}} \to
\mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$, $\nu_b(\emptyset) = 0.8$. Let
$\mu$ be the measure on $2^{2^U}$ induced by $\nu = \nu_a \cdot \nu_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\begin{align*}
  \mu(a) &= \nu(\{ a, b \}) + \nu(\{ a \}) \\
         &= \nu_a(\{a\})\nu_b(\{b\}) + \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
\end{align*}

@@

WMC with weights defined on literals is only able to capture a subset of all
possible measures on the Boolean algebra. This can be illustrated with a simple
example.

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as:\footnote{The value of $\mu$ on any
    other element of the Boolean algebra can be deduced using the definition.}
  \begin{align*}
    \mu(a \land b) &= 0.72, \\
    \mu(a \land \neg b) &= 0.18, \\
    \mu(\neg a \land b) &= 0.07, \\
    \mu(\neg a \land \neg b) &= 0.03.
  \end{align*}
  If $\mu$ could be represented using traditional (factored) WMC, we would have
  to find two weight functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$
  and $\nu_b\colon 2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a
  \cdot \nu_b$ induces $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy
  this system of equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

@@

Since many measures of interests may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\cite{DBLP:journals/ai/ChaviraD08}. We can indeed show that this is always
possible, however, as ensuing sections will demonstrate, it often makes the
inference task much harder.

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables to maintain literal independence. In line with the names of previous
encodings, we shall call this encoding \texttt{db21}.

@@

\begin{algorithm}
  \caption{Encoding a Bayesian network}
  \label{alg:encoding}
  \KwData{vertices $\mathcal{V}$, probability distribution $\Pr$}
  \KwResult{$\phi\colon 2^U \to \mathbb{R}_{\ge 0}$}
  $\phi \gets 1$\;
  \For{$X \in \mathcal{V}$}{
    \textit{let} $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n \}$\;
    $\mathrm{CPT}_X \gets 0$\;
    \eIf{$|\im X| = 2$}{
      \For{$(y_i)_{i=1}^n \in \prod_{i = 1}^n \im Y_i$}{
        $p_1 \gets \Pr(X = 1 \mid y_1, \dots, y_n)$\;
        $p_0 \gets \Pr(X \ne 1 \mid y_1, \dots, y_n)$\;
        \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X$\;
        \pushline $+ p_1[\lambda_{X=1}] \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
        \dosemic $+ p_0 \overline{[\lambda_{X=1}]} \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
      }
    }{
      \textit{let} $\im X = \{ x_1, \dots, x_m \}$\;
      \For{$x \in \im X$ {\rm \textbf{and}} $(y_i)_{i=1}^n \in \prod_{i = 1}^n
        \im Y_i$}{
        $p_x \gets \Pr(X = x \mid y_1, \dots, y_n)$\;
        \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X$\;
        \pushline $+ p_x[\lambda_{X=x}] \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
        \dosemic $+ \overline{[\lambda_{X=x}]} \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
      }
      \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X \cdot \left( \sum_{i=1}^m
        [\lambda_{X = x_i}] \right)$\;
      \pushline\dosemic $\cdot \prod_{i=1}^m \prod_{j=i+1}^m
      (\overline{[\lambda_{X = x_i}]} + \overline{[\lambda_{X = x_j}]})$\;
    }
    $\phi \gets \phi \cdot \mathrm{CPT}_X$\;
  }
  \Return{$\phi$}\;
\end{algorithm}

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\cite{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{It can just as well compute any probability expressed using the
random variables in $\mathcal{V}$, but we focus mostly on marginal probabilities
in our experiments.} First, \cref{lemma:cpt} shows that any conditional weight function
produces the right answer when given an appropriate encoding of
variable-value assignments for a conditional probability in the Bayesian
network. Then, \cref{lemma:full_distribution} shows that $\phi$ represents the
full probability distribution of the Bayesian network, i.e., it gives the right
probabilities for the right inputs and zero otherwise. We end with
\cref{thm:correctness} that shows how $\phi$ can be combined with an encoding of
a single variable-value assignment so that ADDMC would compute its marginal
probability.

@@

@@

@@

\Cref{alg:encoding} encodes a Bayesian network into a function on a Boolean
algebra, but how does it relate to the standard interpretation of a WMC encoding
as a formula in conjunctive normal form (CNF) together with a collection of
weights? The factors of $\phi$ that restrict the values of indicator variables
for non-binary random variables are already expressed as a product of sums of
0/1-valued functions, i.e., a kind of CNF. Disregarding these functions, each
conditional weight function $\mathrm{CPT}_X$ is represented by a sum with a term
for every subset of $\mathcal{E}^*(X)$. To encode these terms, we introduce
\emph{extended weight clauses} to the WMC format used by Cachet
\cite{DBLP:conf/sat/SangBBKP04}. For instance, here is a representation of the
Bayesian network from \cref{fig:example_bn}:
\[
  \begin{array}{lrrll}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &-\lambda\sb{W=1} &0.6 &1 \\
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1 \\
    w &\lambda\sb{T=h} &-\lambda\sb{W=1} &0.1 &1
  \end{array}
\]
where each indicator variable is eventually replaced with a unique positive
integer. Each line prefixed with a $w$ can be split into four parts: the `main'
variable (always not negated), conditions (possibly none), and two weights. For
example, the line
\[
  \begin{array}{lrrll}
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1
  \end{array}
\]
encodes the function $0.3[\lambda_{T=m}] \cdot \overline{[\lambda_{W=1}]} +
1\overline{[\lambda_{T=m}]} \cdot \overline{[\lambda_{W=1}]}$ and can be
interpreted as defining two conditional weights: $\nu(T = m \mid W = 0) = 0.3$,
and $\nu(T \ne m \mid W = 0) = 1$, the former of which corresponds to a row in
the CPT of $T$ while the latter is artificially added as part of the encoding.
In our encoding of Bayesian networks, it is always the case that, in each weight
clause, either both weights sum to one, or the second weight is equal to one.
Finally, note that (without any additional restrictions), the measure induced by
these weight functions is not probabilistic (i.e., $\mu(\top)$ may not be equal
to one).

@@

ADDMC constructs the \emph{Gaifman graph} \cite{gaifman1982local} of the input
CNF formula as an aid for the algorithm's heuristics. This graph has as vertices
the variables of the formula, and there is an edge between two variables $u$ and
$v$ if there is a clause in the formula that contains both $u$ and $v$. We
extend this definition to functions on Boolean algebras, i.e., the factors of
$\phi$. For any pair of distinct variables $u, v \in U$, we draw an edge between
them in the Gaifman graph if there is a function $\alpha\colon 2^X \to
\mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v \in
X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges between
all distinct pairs of variables in $\mathcal{E}^*(X)$.

>>>>

\begin{abstract} % TODO
%1. Introduction. In one sentence, what’s the topic?
  % WMC
%2. State the problem you tackle.
  % Not using additional literals
%3. Summarize (in one sentence) why nobody else has adequately answered the research question yet.
  % The dominant definition of WMC severely restricts what measures can be represented
%4. Explain, in one sentence, how you tackled the research question.
  % Defining weights akin to conditional probabilities.
%5. In one sentence, how did you go about doing the research that follows from your big idea.
  % ADDMC + Bayesian networks
%6. As a single sentence, what’s the key impact of your research?
  % Big improvements in inference speed that can be applied to other algorithms
  % and application domains.
\end{abstract}

@@

Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05} has emerged
as the unifying inference mechanism across many (probabilistic) domains. It has
been applied to Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906} and probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, probabilistic
logic programming languages \cite{DBLP:journals/tplp/FierensBRSGTJR15} and
probabilistic programs with discrete probability distributions
\cite{DBLP:journals/corr/abs-2005-09089}. Over the last fifteen years, WMC has
been extended and generalised in many ways, e.g., to handle continuous
probability distributions \cite{DBLP:conf/ijcai/BellePB15}, first-order
probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, the idea of weights in WMC
has been generalised from non-negative real numbers to an arbitrary semiring,
making it applicable to domains such as parsing, dynamic programming, constraint
programming, databases, and networks \cite{DBLP:journals/japll/KimmigBR17}.

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This simplification is motivated by the fact that the
number of models scales exponentially with the number of atoms, so listing the
weight of every model is intractable. However, this also severely restricts what
probability distributions can be represented. A common way to overcome this
limitation is by adding more literals. While we show that this is always
possible in \cref{sec:wmc_as_measure}, the rest of the paper demonstrates that
it can be significantly more efficient to encode weights in a more flexible
format instead.

% TODO: highlight the main strengths

@@

In this section, we give a brief introduction to two alternative ways to think
about logical constructs such as models and formulas. Let $\mathcal{L}$ be a
propositional logic with atoms $a$ and $b$, and let $U = \{ a, b \}$. Then
$2^U$, the power set of $U$, is the set of all models of $\mathcal{L}$, and
$2^{2^U}$ is the set of all formulas. These sets can also be represented as
Boolean algebras (e.g., using the syntax $(2^{2^U}, \land, \lor, \neg, \bot,
\top)$) with a partial order $\le$ that corresponds to set inclusion
$\subseteq$---see \cref{tbl:notation_example} for examples of how various
elements can be represented in both notations. Most importantly, note that the
word \emph{atom} has completely different meanings in logic and in Boolean
algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an element of
$U$, whereas an atom in a Boolean algebra is (in set-theoretic terms) a
singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom and refer to logical
atoms as \emph{variables}. In the rest of the paper, for any set $U$, we will
use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

\subsection{The Space of Functions on Boolean Algebras}
% TODO: clarify why this is needed

We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. All of
these operations have efficient implementations in the CUDD
\cite{somenzi1998cudd} package for manipulating ADDs (among other things) that
is used by ADDMC \cite{DBLP:conf/aaai/DudekPV20}. The definitions of
multiplication and projection are as defined by \citet{DBLP:conf/aaai/DudekPV20}
while others are new.

@@

\section{WMC as a Measure on a Boolean Algebra} \label{sec:wmc_as_measure}

In this section, we introduce an alternative definition of WMC and demonstrate
how it relates to the standard one. Let $U$ be a set. A \emph{measure} is a
function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$,
and $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a
\land b = \bot$ \cite{gaifman1964concerning,DBLP:books/daglib/0090259}. A
\emph{weight function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$. A
weight function is \emph{factored} if $\nu = \prod_{x \in U} \nu_x$ for some
functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that
a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a
measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if $\mu_\nu(x) =
\sum_{\{u\} \le x} \nu(u)$.

@@

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a weight function defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \begin{split}
    \mathrm{WMC}(\Delta) &= \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) \\
    &= w(a)w(b) + w(a)w(\neg b) = 0.3.
  \end{split}
\end{equation}
Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon 2^{\{b\}} \to
\mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$, $\nu_b(\emptyset) = 0.8$. Let
$\mu$ be the measure on $2^{2^U}$ induced by $\nu = \nu_a \cdot \nu_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\begin{align*}
  \mu(a) &= \nu(\{ a, b \}) + \nu(\{ a \}) \\
         &= \nu_a(\{a\})\nu_b(\{b\}) + \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
\end{align*}
Thus, one can equivalently think of WMC as summing over models of a theory or
over atoms below an element of a Boolean algebra.

@@

Using this new definition of WMC, we can show that WMC with weights defined on
literals is only able to capture a subset of all possible measures on a
Boolean algebra. This can be demonstrated with a simple example.

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as:\footnote{The value of $\mu$ on any
    other element of $2^{2^U}$ can be deduced from the definition of a measure.}
  \begin{align*}
    \mu(a \land b) &= 0.72, \\
    \mu(a \land \neg b) &= 0.18, \\
    \mu(\neg a \land b) &= 0.07, \\
    \mu(\neg a \land \neg b) &= 0.03.
  \end{align*}
  If $\mu$ could be represented using literal-weight (factored) WMC, we would
  have to find two weight functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge
    0}$ and $\nu_b\colon 2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu =
  \nu_a \cdot \nu_b$ induces $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to
  satisfy this system of equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\cite{DBLP:journals/ai/ChaviraD08}. We can indeed show that this is always
possible, however, as ensuing sections will demonstrate, it often makes the
inference task much harder.

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables. In line with the names of the previous encodings, we shall call this
encoding \texttt{db21}.

@@

\begin{algorithm}[t]
  \caption{Encoding a Bayesian network}
  \label{alg:encoding}
  \KwData{vertices $\mathcal{V}$, probability distribution $\Pr$}
  \KwResult{$\phi\colon 2^U \to \mathbb{R}_{\ge 0}$}
  $\phi \gets 1$\;
  \For{$X \in \mathcal{V}$}{
    \textit{let} $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n \}$\;
    $\mathrm{CPT}_X \gets 0$\;
    \eIf{$|\im X| = 2$}{
      \For{$(y_i)_{i=1}^n \in \prod_{i = 1}^n \im Y_i$}{
        $p_1 \gets \Pr(X = 1 \mid y_1, \dots, y_n)$\;
        $p_0 \gets \Pr(X \ne 1 \mid y_1, \dots, y_n)$\;
        \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X$\;
        \pushline $+ p_1[\lambda_{X=1}] \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
        \dosemic $+ p_0 \overline{[\lambda_{X=1}]} \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
      }
    }{
      \textit{let} $\im X = \{ x_1, \dots, x_m \}$\;
      \For{$x \in \im X$ {\rm \textbf{and}} $(y_i)_{i=1}^n \in \prod_{i = 1}^n
        \im Y_i$}{
        $p_x \gets \Pr(X = x \mid y_1, \dots, y_n)$\;
        \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X$\;
        \pushline $+ p_x[\lambda_{X=x}] \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
        \dosemic $+ \overline{[\lambda_{X=x}]} \cdot \prod_{i=1}^n
        [\lambda_{Y_i=y_i}]$\;
      }
      \nosemic $\mathrm{CPT}_X \gets \mathrm{CPT}_X \cdot \left( \sum_{i=1}^m
        [\lambda_{X = x_i}] \right)$\;
      \pushline\dosemic $\cdot \prod_{i=1}^m \prod_{j=i+1}^m
      (\overline{[\lambda_{X = x_i}]} + \overline{[\lambda_{X = x_j}]})$\;
    }
    $\phi \gets \phi \cdot \mathrm{CPT}_X$\;
  }
  \Return{$\phi$}\;
\end{algorithm}

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\cite{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{It can just as well compute \emph{any} probability expressed
  using the random variables in $\mathcal{V}$. We only focus on marginal
  probabilities for notational simplicity.} We begin with \cref{lemma:cpt} which
shows that any conditional weight function produces the right answer when given
a valid encoding of variable-value assignments relevant to the CPT.

@@

Now, \cref{lemma:full_distribution} shows that $\phi$ represents the full
probability distribution of the Bayesian network, i.e., it gives the right
probabilities for the right inputs and zero otherwise.

@@

We end with \cref{thm:correctness} that shows how $\phi$ can be combined with an
encoding of a single variable-value assignment so that ADDMC would compute its
marginal probability.

@@

\Cref{alg:encoding} encodes a Bayesian network into a function on a Boolean
algebra, but how does it relate to the standard interpretation of a WMC encoding
as a formula in conjunctive normal form (CNF) together with a collection of
weights? The factors of $\phi$ that restrict the values of indicator variables
for non-binary random variables are already expressed as a product of sums of
0/1-valued functions, i.e., a kind of CNF. Disregarding these functions, each
conditional weight function $\mathrm{CPT}_X$ is represented by a sum with a term
for every subset of $\mathcal{E}^*(X)$. To encode these terms, we introduce
\emph{extended weight clauses} to the WMC format used by Cachet
\cite{DBLP:conf/sat/SangBBKP04}. For instance, here is a representation of the
Bayesian network from \cref{fig:example_bn}:
\[
  \begin{array}{lrrll}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &-\lambda\sb{W=1} &0.6 &1 \\
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1 \\
    w &\lambda\sb{T=h} &-\lambda\sb{W=1} &0.1 &1
  \end{array}
\]
where each indicator variable is eventually replaced with a unique positive
integer. Each line prefixed with a $w$ can be split into four parts: the `main'
variable (always not negated), conditions (possibly none), and two weights. For
example, the line
\[
  \begin{array}{lrrll}
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1
  \end{array}
\]
encodes the function $0.3[\lambda_{T=m}] \cdot \overline{[\lambda_{W=1}]} +
1\overline{[\lambda_{T=m}]} \cdot \overline{[\lambda_{W=1}]}$ and can be
interpreted as defining two conditional weights: $\nu(T = m \mid W = 0) = 0.3$,
and $\nu(T \ne m \mid W = 0) = 1$, the former of which corresponds to a row in
the CPT of $T$ while the latter is artificially added as part of the encoding.
In our encoding of Bayesian networks, it is always the case that, in each weight
clause, either both weights sum to one, or the second weight is equal to one.
Finally, note that the measure induced by these weights is not probabilistic
(i.e., $\mu(\top) \ne 1$) by itself, but it becomes probabilistic when combined
with the additional clauses that restrict what combinations of indicator
variables can co-occur.

@@

Here we describe two changes to
ADDMC\footnote{\url{https://github.com/vardigroup/ADDMC}}
\cite{DBLP:conf/aaai/DudekPV20} needed to adapt it to the new format.

First, ADDMC constructs the \emph{Gaifman graph} \cite{gaifman1982local} of the
input CNF formula as an aid for the algorithm's heuristics. This graph has as
vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the Gaifman graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v
\in X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges
between all distinct pairs of variables in $\mathcal{E}^*(X)$.

Second, even though the function $\phi$ produced by \cref{alg:encoding} is
constructed to have $2^U$ as its domain, sometimes the domain is effectively
reduced to $2^V$ for some $V \subset U$ by the ADD manipulation algorithms that
optimise the ADD representation of a function. For a simple example, consider
$\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) =
\alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to $\alpha'\colon
2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as $\alpha'(\emptyset) = 0.5$. To
compensate for these reductions, for the original WMC format with a weight
function $w\colon U \cup \{ \neg u \mid u \in U \} \to \mathbb{R}_{\ge 0}$,
ADDMC would multiply its computed answer by $\prod_{u \in U \setminus V} w(u) +
w(\neg u)$. With the new WMC format, we instead multiply the answer by $2^{|U
  \setminus V|}$. Each `excluded' variable $u \in U \setminus V$ satisfies two
properties: all weights associated with $u$ are equal to $0.5$ (otherwise the
corresponding CPT would depend on $u$, and $u$ would not be excluded), and all
other CPTs are independent of $u$ (or they may have a trivial dependence, where
the probability stays the same if $u$ is replaced with its complement). Thus,
the CPT that corresponds to $u$ still multiplies the weight of every atom in the
Boolean algebra by $0.5$, but the number of atoms being considered is halved. To
correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.
